---
title: "HW 3"
author: "Jared Brotamonte"
date: "9/22/2023"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Exercise 1
```{r}
library(ISLR2)
data("Auto")
library(ggplot2)
```


## (a)
```{r}
# Perform simple linear regression
model <- lm(mpg ~ horsepower, data = Auto)

# Print the summary of the regression results
summary(model)
```
i. Yes, there is a relationship between the predictor "horsepower" and the response "mpg" given the p-value is "<2e-16"

ii. The multiple R-squared value is 0.6059 thus this means that about 60.59% of the variance in "mpg" can be explained by the "horsepower" predictor. This indicates a moderate to strong relationship between the two.

iii. The coefficient for "horsepower" is -0.157845. Since this is less than zero, it indicates a negative relationship between "horsepower" and "mpg" or in otherwords as "horsepower" increases by 1, "mpg" should decrease by about -0.157845.

## (b)
```{r}
new_data <- data.frame(horsepower = 98)

# Use the 'predict' function to make the prediction
predicted_mpg <- predict(model, newdata = new_data)

# Print the predicted 'mpg' value
print(predicted_mpg)
```

## (c)
```{r}
# Create a scatterplot of mpg vs. horsepower
plot(Auto$horsepower, Auto$mpg, xlab = "Horsepower", ylab = "MPG", main = "Scatterplot of MPG vs. Horsepower")

# Add the least squares regression line
abline(model, col = "red")
```

## (d)
```{r}
# Create diagnostic plots for the linear regression model
par(mfrow = c(2, 2))  # Set up a 2x2 grid for the plots
plot(model)
```
Looking at the residuals vs fitted graph, there is a slight curve to the graph thus meaning a more complex model might be better for the data.

# Exersize 2
```{r}
set.seed(1)
```

## (a)
```{r}
X <- rnorm(100, mean = 0, sd = 1)
```

## (b)
```{r}
epsilon <- rnorm(100, mean = 0, sd = 0.25)
```

## (c)
```{r}
# Calculate Y based on the linear model
Y <- -1 + 0.5 * X + epsilon

# (i) Length of vector Y
length_Y <- length(Y)

# (ii) Values of beta
beta_0 <- -1
beta_1 <- 0.5
```
i.
```{r}
length_Y
```
ii.
```{r}
beta_0
beta_1
```

## (d)
```{r}
# Create a scatterplot of X vs. Y
plot(X, Y, main = "Scatterplot of X vs. Y", xlab = "X", ylab = "Y")
```
Based off the plot above, there seems to be a positive realationship between X and Y, and the pattern is rather linear.

## (e)
```{r}
# Fit a least squares linear model
model <- lm(Y ~ X)

# Print the summary of the model
summary(model)
```

The actual beta 0 was estimated at about -0.9745 while the actual beta 1 was estimated at about 0.5213. This is extremely close to the predicted beta values from earlier. Another thing to note from this model is that p-value for the relationship between x and y is extremely small thus meaning significance.